#Reproducible Research
-Peer Assessment 1 - Peter Polack

##Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected and include the number of steps taken in 5 minute intervals each day.

##Assignment
This assignment will be described in multiple parts. You will need to write a report that answers the questions detailed. Ultimately, you will need to complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file.

Throughout your report make sure you always include the code that you used to generate the output you present.

###Loading and preprocessing the data
Read in working file, assumes working file 'activity.csv' is present in the workding directory
```{r}
     activity <- read.csv("activity.csv")
```
     
###  Create initial dataframes
 - dailysteps - total amount of steps by day
 - intervalsteps - average amount of steps by time interval
```{r}
     dailysteps <- aggregate(list(steps=activity$steps), 
          by = list(date = activity$date), FUN = "sum", na.rm=TRUE)
     
     intervalsteps <- aggregate(list(steps=activity$steps), 
          by = list(interval = activity$interval), FUN = "mean", na.rm=TRUE)
```
 
## What is mean total number of steps taken per day?
### Mean number of steps per day
```{r}
     originalmean <- mean(dailysteps$steps, na.rm=TRUE)
     originalmean
```
### Median number of steps per day
```{r}
     originalmedian <- median(dailysteps$steps, na.rm=TRUE)
     originalmedian
```
### Histogram of number of steps per day
```{r}
     hist(dailysteps$steps, xlab = "# Daily Steps", 
          main="Frequency of # of Daily Steps")
```


## What is the average daily activity pattern?
```{r}
     plot(intervalsteps$interval, intervalsteps$steps, type="l", 
          xlab="Minute intervals", ylab="Steps (mean)")
```

### Which 5-minute interval, on average, contains the maximum number of steps?
```{r}
     intervalsteps[which.max(intervalsteps[,2]),]
```

  
    
## Inputing missing values

### How many NAs are in the original dataset?
```{r}
     sapply(activity, function(x) sum(is.na(x)))
```

### Create copy of original data set and fill NAs with alternative value (explained below)
 - Fill copied dataset NAs with averaged interval steps figured out previously in assignment.
 - For example: interval 0 NA would be filled with interval 0 average steps.
 
```{r}
     filledactivity <- activity
     filledactivity[is.na(filledactivity)] = intervalsteps$step
```

### Create new "dailyfilledsteps" aggregate, then create new mean, median and histogram based on them
```{r}
     dailyfilledsteps <- aggregate(list(steps=filledactivity$steps), 
          by = list(date = filledactivity$date), FUN = "sum", na.rm=TRUE)
```

### Original and new mean number of steps per day, based on updated data
```{r}
     originalmean
     newmean <- mean(dailyfilledsteps$steps, na.rm=TRUE)
     newmean
```

### Original and new median number of steps per day, based on updated data
```{r}
     originalmedian
     newmedian <- median(dailyfilledsteps$steps, na.rm=TRUE)
     newmedian
```

### Histogram of number of steps per day, based on updated data
```{r}
     hist(dailyfilledsteps$steps, xlab = "# Daily Steps", 
          main="Frequency of # of Daily Steps")
```

 - New mean and median are higher than original.
 - New histogram is more bell curved.
 - Both of these results are due to replacing the NAs with the average steps per interval.  
 
## Are there differences in activity patterns between weekdays and weekends?
 - Use updated copy of the original activity dataset
 - Add weekday column that contains the day of the week for that row's date

```{r}
     filledactivity$weekday <- weekdays(as.Date(as.character(filledactivity$date)))
```

 - Add weekend column that contains "weekday"/"weekend" factor variable
```{r}
     filledactivity$weekend <- ifelse((filledactivity$weekday=="Saturday"|
          filledactivity$weekday=="Sunday"),"weekend", "weekday")
     filledactivity$weekend <- as.factor(filledactivity$weekend)
```

 - Create new aggregate of interval steps on the weekend factor column and interval
```{r}
     filledintervalsteps <- aggregate(list(steps=filledactivity$steps), by = list(weekend = filledactivity$weekend, interval = filledactivity$interval), FUN = "mean", na.rm=TRUE)

```

### Plot out interval steps by weekend and weekday
 - It seems like there was more steps in the morning on weekdays (perhaps people were getting workouts in the mornings)  
 - Step distribution was more even over the weekends, perhaps based on more general activity during a weekend

```{r}
     library(lattice)
     xyplot(filledintervalsteps$steps ~ filledintervalsteps$interval | 
          filledintervalsteps$weekend, layout = c(1, 2), type="l")
```